---
title: "GSEA"
output:
  pdf_document: default
  html_document: default
---
Manual for fgsea:
http://bioconductor.org/packages/release/bioc/manuals/fgsea/man/fgsea.pdf
-> Further information on how to create for example enrichment plots

```{r}
######################
#######USER INPUT####
######################
#Data to import gene list
FILENAME = "dge_presults.csv"
FILEPATH = paste("./",FILENAME, sep = "")

#Data to export final files 
LISTNAME = "presults" #filename to append to saved files
######################
######################
```
#1. Import required libraries
```{r}
# Import required libraries
library(GSA, quietly=TRUE, warn.conflicts=FALSE, verbose=FALSE)       #To read in gene sets in gmt format
library(rmarkdown, quietly=TRUE, warn.conflicts=FALSE, verbose=FALSE)
library(ggdendro,quietly=TRUE,warn.conflicts=FALSE,verbose=FALSE)
library(dendextend,quietly=TRUE,warn.conflicts=FALSE,verbose=FALSE)
library(ggplot2, quietly=TRUE, warn.conflicts=FALSE, verbose=FALSE)
library(colorspace, quietly=TRUE, warn.conflicts=FALSE, verbose=FALSE)
source("https://bioconductor.org/biocLite.R")
biocLite("fgsea") #Package that runs GSEA
biocLite("biomaRt")
library(biomaRt, quietly=TRUE, warn.conflicts = FALSE, verbose = FALSE)
setwd("/Users/esidhom/Dropbox (MIT)/Harvard_MD-PhD/GrekaLab/Experiments/Analysis")

# Create mouse and human biomaRts
human <- useMart(host="www.ensembl.org", "ENSEMBL_MART_ENSEMBL", dataset="hsapiens_gene_ensembl")
mouse <- useMart(host="www.ensembl.org", "ENSEMBL_MART_ENSEMBL", dataset="mmusculus_gene_ensembl")
```
#2. Download gene sets from MSig database in gmt format on your local machine
#3. Load gene sets in R
```{r}
#Example how to load gene sets into R (here: KEGG)
GOpathways=fgsea::gmtPathways("../GeneSets/c5.bp.v6.2.symbols.gmt")
KEGGpathways=fgsea::gmtPathways("../GeneSets/c2.cp.kegg.v6.2.symbols.gmt")
```
#4. You need a ranked file of your genes of interest. Input is always all genes! Rank them for example by p-value: rank = -log10(pval)*sign(FC)
```{r}
#ranklist by -log10(pvalue)*SIGN(log2fc)
genelist <- read.table(FILEPATH,header = TRUE, sep = " ")
if (min(genelist$p_val) == 0) {new_min = unique(genelist$Pvalue)[2]/1e10
genelist$p_val[genelist$p_val == 0] <- new_min}
ranklist <- data.frame(row.names(genelist),-log(genelist$Pvalue)*sign(genelist$log2FC))
colnames(ranklist) <- c("geneNames","rank")
ranklist <- ranklist[order(ranklist$rank, decreasing = TRUE),]
ranklist <- lapply(ranklist, function(x) if (is.factor(x)) as.character(x) else {x})

#ranklist by zscore
ranklist2 <- data.frame(row.names(genelist),genelist$Stat)
colnames(ranklist2) <- c("geneNames","rank")
ranklist2 <- ranklist2[order(ranklist2$rank, decreasing = TRUE),]
ranklist2 <- lapply(ranklist2, function(x) if (is.factor(x)) as.character(x) else {x})

#The ranked lists (2 columns: Gene, rank) has to be converted to a named vector (vector for which each entry is assigned to an attribute. Here: The rank-value is assigned to the gene name)
rnk_file = setNames(ranklist$rank, ranklist$geneNames)
rnk_file2 = setNames(ranklist2$rank, ranklist2$geneNames)
```
#5. Convert mouse to human gene names
```{r}
# Get human gene names
annotation <- getLDS(attributes = c("mgi_symbol"),
                     filters = "mgi_symbol", values = names(rnk_file2),
                     mart = mouse,
                     attributesL = c("hgnc_symbol"), martL = human) 

# Make the rownames of res into a data.frame
mouseRows <- data.frame(ranklist2$geneNames)
# Merge mouse and human gene names
annotationMerged <- merge(x=mouseRows,
                          y=annotation,
                          by.x="ranklist2.geneNames",
                          by.y="MGI.symbol",
                          all=TRUE)
# Order mouseRownames to match rownames of dds_2weeks
## We don't want to remove "NA" here. We need to keep all rows in dds_2weeks
annotationOrdered <- annotationMerged[match(ranklist2$geneNames, annotationMerged$ranklist2.geneNames),]
annotationOrdered[is.na(annotationOrdered)] <- "NA"
# Check if all rownames annotationOrdered equal rownames of dds_2weeks
all(ranklist2$geneNames == annotationOrdered$ranklist2.geneNames) 
# Make human gene names a sepamousee data.frame with mouse gene names as row names
humanGeneNames <- annotationOrdered$HGNC.symbol
# Add humanGeneNames to dds_2weeks
rnk_file_hum2 = setNames(ranklist2$rank, humanGeneNames)
```
#6. Run GSEA
-> Output: GSEA statistics with ES, NES, pval, padj,leading edge

```{r}
#Input: GeneSet, ranked file, number of permutations, minSize of GeneSet, maxSize of GeneSet
#Output: table of pathways and pdf with enrichment plots

#define GSEA function
GSEA_run <- function(Pathways,Stats,NAME,LISTNAME){
  #run GSEA
  print(paste("Running GSEA: ", NAME, sep=""))
  GSEA_object = fgsea::fgsea(Pathways,Stats,nperm=10000, minSize=15, maxSize=500, nproc=0, gseaParam=1,BPPARAM=NULL)
  
  #only proceed with downstream steps if pathways found
  if (nrow(GSEA_object) >0) {
    #make GSEA leading edge column into simple list of genes rather than vectors
    GSEA_object$leadingEdge <- gsub("[^0-9A-Z ]", "", paste(GSEA_object$leadingEdge))
    
    #order table by increasing adjusted p-value and write pathway table to file
    GSEA_object <- GSEA_object[order(GSEA_object$padj, decreasing = FALSE),]
    
    write.table(GSEA_object,file = paste("GSEA/", format(Sys.time(), "%Y%b%d_%H%M"),"_GSEA_", LISTNAME, "_", NAME, ".csv", sep = ""), sep=",", row.names = FALSE)
    
    #make pdf of enrichment plots
    FILENAME <- paste(format(Sys.time(), "%Y%b%d_%H%M"),"_GSEA_", LISTNAME, "_", NAME, ".pdf", sep = "")
    FILEPATH <- paste("./figures/gsea/",FILENAME,sep="")
    pdf(file=FILEPATH,useDingbats=FALSE)
    if (nrow(GSEA_object) <= 50) { for (i in 1:nrow(GSEA_object)) {p <- fgsea::plotEnrichment(Pathways[[GSEA_object$pathway[i]]],Stats) + ggtitle(GSEA_object$pathway[i])
    print(p)}} else {for (i in 1:50) {p <- fgsea::plotEnrichment(Pathways[[GSEA_object$pathway[i]]],Stats) + ggtitle(GSEA_object$pathway[i])
    print(p)}}
    dev.off()
  } else {print(paste("No ", NAME, " pathways found.", sep=""))}
  
  return(GSEA_object)
}

#run GSEA with different genesets
GSEA_KEGG <- GSEA_run(KEGGpathways,rnk_file_hum2,"KEGG",LISTNAME)
#GSEA_GO <- GSEA_run(GOpathways,rnk_file_hum,"GO",LISTNAME)

```
#7. Leading edge analysis

```{r}
GSEA_LEanalysis <- function(GSEA_object, geneList, annotation, LISTNAME, NAME, k1, k2,TEXTSIZE){
  #this function takes in the GSEA output, and file name/path for saving PDF, the name of the pathway database used, and number of clusters to sub-divide dendrograms (k1, k2). It plots up- & down-regulated pathways separately. 
  
  #split GSEA object into up- and down-regulated pathways
  GSEA_object <- GSEA_object[order(GSEA_object$NES, decreasing = TRUE),]
  GSEA_up <- GSEA_object[which(GSEA_object$NES > 0),]
  GSEA_down <- GSEA_object[which(GSEA_object$NES < 0),]
  
  #perform leading edge analysis on each set separately
  DIRECTION="UP in CTRL"
  FILENAME <- paste(format(Sys.time(), "%Y%b%d_%H%M"),"_LEAnalysis_", LISTNAME, "_", NAME, "_", gsub("[^A-z]","",DIRECTION), ".pdf", sep = "")
  FILEPATH <- paste("figures/gsea/",FILENAME,sep="")
  pdf(file=FILEPATH,useDingbats=FALSE)
  GSEA_LE(GSEA_split=GSEA_up, geneList=geneList, annotation=annotation, LISTNAME=LISTNAME, NAME=NAME, DIRECTION=DIRECTION, k=k1, TEXTSIZE=TEXTSIZE)
  dev.off()
  
  DIRECTION="UP in KDKD"
  FILENAME <- paste(format(Sys.time(), "%Y%b%d_%H%M"),"_LEAnalysis_", LISTNAME, "_", NAME, "_", gsub("[^A-z]","",DIRECTION), ".pdf", sep = "")
  FILEPATH <- paste("figures/gsea/",FILENAME,sep="")
  pdf(file=FILEPATH,useDingbats=FALSE)
  GSEA_LE(GSEA_split=GSEA_down, geneList=geneList, annotation=annotation, LISTNAME=LISTNAME, NAME=NAME, DIRECTION=DIRECTION, k=k2, TEXTSIZE=TEXTSIZE)
  dev.off()
}
GSEA_LE <- function(GSEA_split, geneList, annotation, LISTNAME, NAME, DIRECTION, k, TEXTSIZE){
  
  #if pathway list is >100 pathways, select the top 100 by adj. p-value for efficiency 
  if (nrow(GSEA_split) > 100) {
    GSEA_split <- GSEA_split[order(GSEA_split$padj, decreasing = FALSE),]
    GSEA_split <- GSEA_split[1:100,]
  }
  #create a list holding all leading edge genes for each pathway
  GeneList = list()
  for (i in 1:nrow(GSEA_split)) {
    GeneList[i] <- as.list(strsplit(GSEA_split$leadingEdge[i],"[[:space:]]"))
  }
  names(GeneList) <- GSEA_split$pathway
  
  #create distance matrix of overlap between all genes
  #Overlap = #intersection(A,B)/max(size(A), size(B))
  Overlap.matrix <- matrix(0L, nrow = nrow(GSEA_split), ncol = nrow(GSEA_split))
  for (i in 1:nrow(GSEA_split)) {
    for (j in 1:nrow(GSEA_split)) {
      Intersection = length(intersect(GeneList[i][[GSEA_split$pathway[i]]], GeneList[j][[GSEA_split$pathway[j]]]))
      Size = max(c(length(GeneList[i][[GSEA_split$pathway[i]]]), length(GeneList[j][[GSEA_split$pathway[j]]])))
      Overlap.matrix[i,j] <- Intersection/Size
    }
  }
  
  #calculate distance matrix and make dendrogram. Divide dendrogram in k clusters and plot each separately.
  Overlap.matrix <- data.frame(Overlap.matrix) #convert overlap matrix to a data frame
  rownames(Overlap.matrix) <- GSEA_split$pathway #label rows with pathways
  hc <- hclust(dist(Overlap.matrix, method = "euclidean"), method = "average") #calculate distance matrix
  dend <- as.dendrogram(hc) #convert to dendrogram
  
  cols <- rainbow_hcl(k) #select colors for clusters
  dend <- dend %>% color_branches(k = k, col = cols) %>% 
    set("branches_lwd", 0.5)
  clust    <- cutree(dend,k=k) #cluster dendrogram based on k-means
  clust.df <- data.frame(label=names(clust), cluster=factor(clust)) #create data-frame with pathways & cluster assignments) 
  
  ggd1 <- as.ggdend(dend) #convert to ggplot2 structure 
  
  #plot full dendrogram
  print(ggplot(ggd1, labels = FALSE) + 
          ggtitle(paste(NAME, " - ", DIRECTION, sep=""))+
          coord_flip() + 
          scale_x_continuous(breaks = label(ggd1)$x, 
                             labels = label(ggd1)$label,
                             position = "top") +
          scale_y_reverse(expand = c(0, 0)) +
          theme_minimal() + 
          theme(axis.title = element_blank(),
                axis.text.y = element_text(size = rel(TEXTSIZE)),
                panel.grid = element_blank()))
  
  
  #prune original dendrogram into smaller dendrograms
  dends <- list()
  for(i in 1:k) {
    labels_to_prune <- clust.df$label[which(clust.df$cluster != i)]
    dends[[i]] <- prune(dend, labels_to_prune)
  }
  
  #plot dendrograms of individual clusters and display most highly represented genes
  for(i in 1:k) {
    Clust.labels <- clust.df$label[which(clust.df$cluster == i)]
    if (length(Clust.labels) > 1) {
      temp <- as.ggdend(dends[[i]])
      
      print(ggplot(temp, labels = FALSE) + 
              ggtitle(paste0("TREE ", i)) +
              coord_flip() + 
              scale_x_continuous(breaks = label(temp)$x, 
                                 labels = label(temp)$label,
                                 position = "top") +
              scale_y_reverse(expand = c(0, 0)) +
              theme_minimal() + 
              theme(axis.title = element_blank(),
                    axis.text.y = element_text(size = rel(TEXTSIZE)),
                    panel.grid = element_blank()))
      
      Merge.genes = list()
      for (j in 1:length(Clust.labels)) {
        temp_label <- as.character(Clust.labels[j])
        temp_index <- which(clust.df$label == temp_label)
        Merge.genes = c(Merge.genes,GeneList[temp_index][[temp_label]])
      }
      Merge.genes <- as.character(Merge.genes)
      Unique.genes <- unique(Merge.genes)
      Count.vector <- matrix(0L, nrow = length(Unique.genes), ncol = 1)
      for (j in 1:length(Unique.genes)) {
        temp_gene <- Unique.genes[j]
        Count.vector[j] <- length(which(Merge.genes == temp_gene))
      }
      
      Count.vector <- data.frame(Count.vector,Unique.genes, stringsAsFactors = FALSE)
      Count.vector <- Count.vector[order(Count.vector$Unique.genes),]
      rownames(Count.vector) <- Count.vector$Unique.genes
      Count.vector$Unique.genes <- NULL
      colnames(Count.vector) <- "Count"
      
      annotation$HGNC.symbol <- gsub("[^0-9A-Z ]","",annotation$HGNC.symbol)
      
      mouse.genes <- annotation[is.element(annotation$HGNC.symbol,Unique.genes),]
      mouse.genes$ranklist2.geneNames <- as.character(mouse.genes$ranklist2.geneNames)
      
      if (dim(mouse.genes)[1] > dim(Count.vector)[1]) {
        print("Found duplicate genes")
        countgenes <- as.data.frame(table(mouse.genes$HGNC.symbol))
        multiplegenes <- countgenes$Var1[countgenes$Freq > 1]
        for (m in 1:length(multiplegenes)){
          temp.gene <- multiplegenes[m]
          print(temp.gene)
          temp.msgenes <- which(mouse.genes$HGNC.symbol == temp.gene)
          print(mouse.genes$ranklist2.geneNames[temp.msgenes])
          mouse.genes <- mouse.genes[-temp.msgenes[2:length(temp.msgenes)],]
        }
      }
      rownames(mouse.genes) <- mouse.genes$HGNC.symbol
      mouse.genes$HGNC.symbol <- NULL
      colnames(mouse.genes) <- "Gene"
      
      Gene.counts <- as.data.frame(geneList[is.element(rownames(geneList),mouse.genes$Gene),])
      Gene.counts$pct.1 <- NULL
      Gene.counts$pct.2 <- NULL
      
      Gene.counts <- cbind(mouse.genes,Gene.counts)
      Gene.counts <- Gene.counts[order(rownames(Gene.counts)),]
      Gene.counts <- cbind(Gene.counts,Count.vector)
      
      Gene.counts <- Gene.counts[order(Gene.counts$Pvalue),]
      Gene.counts <- Gene.counts[order(Gene.counts$Count, decreasing = T),]
      
      FILENAME <- paste(format(Sys.time(), "%Y%b%d_%H%M"),"_LEAnalysisGeneList_", LISTNAME, "_", NAME, "_", gsub("[^A-z]","",DIRECTION),"_Cluster#_", i, ".csv", sep = "")
      FILEPATH <- paste("GSEA/",FILENAME,sep="")
      write.table(Gene.counts, file=FILEPATH, sep=",",row.names = TRUE)
      
      print(paste(NAME, "pathways, ", DIRECTION, ", Cluster #", i, " - most significant genes:", sep="" ))
      print.data.frame(Gene.counts[1:10,])
    }
    
  }
  
}

GSEA_LEanalysis(GSEA_KEGG, genelist, annotationOrdered, LISTNAME, NAME="KEGG", k1=3, k2=3,TEXTSIZE=1)
#GSEA_LEanalysis(GSEA_GO, genelist, annotation,Ordered, LISTNAME, NAME="GO", k1=3, k2=3,TEXTSIZE=0.8)

```


